---
title: "LIFE EXPECTANCY ANALYSIS"
author: "MEGHANA ATLURI"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  tidy = TRUE,
  comment = "#>",
  fig.align = 'center',
  echo = TRUE
)
```

```{r}
# importing library
library(tidyverse)
library(dplyr)
library(ggplot2)
library(janitor)
library(lubridate)
library(gghighlight)
```

### [Data]{.underline}

```{r}
data <- read_csv("Life Expectancy Data.csv")
```

```{r}
#Dimensions
dim(data)
```

```{r}
#Glimpse
glimpse(data)
```

```{r}
#Columns in data
colnames(data)
```

### [Data Cleaning]{.underline}

```{r}
# Rename columns 
data <- data |>
  clean_names() |>
  rename(
    country = country,
    year = year,
    status = status,
    life_expectancy = life_expectancy,
    adult_mortality = adult_mortality,
    infant_deaths = infant_deaths,
    alcohol = alcohol,
    perc_expenditure = percentage_expenditure,
    hepatitis_b = hepatitis_b,
    measles = measles,
    bmi = bmi,
    under5_deaths = under_five_deaths,
    polio = polio,
    total_expenditure = total_expenditure,
    diphtheria = diphtheria,
    hiv_aids = hiv_aids,
    gdp = gdp,
    population = population,
    thinness_1_19 = thinness_1_19_years,
    thinness_5_9 = thinness_5_9_years,
    income_comp = income_composition_of_resources,
    schooling = schooling
  )

```

```{r}
#Drop Columns
data <- select(data, -c(12,19,20))
```

```{r}
#Check NAs
count_na <- function(x) {
  sum(is.na(x))
}
apply(data, MARGIN = 2, FUN = count_na)
```

```{r}
#Drop rows with missing life_expectancy (target variable)
data <- data |> filter(!is.na(life_expectancy))
```

```{r}
#Drop NAs
data <- data |> drop_na()
```

```{r}
#Converting DataTypes
data <- data |>
  mutate(
    year = as.factor(year),
    status = as.factor(status),
    country = as.factor(country)
  )
```

## Exploratory Data Analysis

```{r}
summary(data)
```

```{r}
#Distribution of Life Expectancy
ggplot(data, aes(x = life_expectancy)) +
  geom_histogram(fill = "steelblue", bins = 30, color = "white") +
  theme_minimal() +
  labs(title = "Distribution of Life Expectancy", 
       x = "Life Expectancy", 
       y = "Count")
```

While many countries enjoy life expectancy above 70 years, a notable proportion still falls below 60yrs. The distribution is **right-skewed.**

```{r}
#Life Expectancy vs Alcohol Consumption
ggplot(data, aes(x = alcohol, y = life_expectancy, color = alcohol)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_color_gradient(low = "#0091ff", high = "#f0650e") +
  theme_minimal() +
  labs(title = "Life Expectancy vs Alcohol Consumption",
       x = "Alcohol (litres/year)",
       y = "Life Expectancy")

```

While the scatter plot shows a **positive association** between alcohol consumption and life expectancy, It is more likely that both variables are influenced by underlying factors — particularly national income level, which affects both healthcare access and lifestyle.

**Correlation is not Causation-** Just because they are correlated doesn't mean that one causes the other.

```{r}
#Life Expectancy by Status
ggplot(data, aes(x = status, y = life_expectancy, fill = status)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Life Expectancy by Status",
       x = "Country Status",
       y = "Life Expectancy")
```

The boxplot clearly illustrates that developed countries tend to have significantly higher and more consistent life expectancy compared to developing countries.

This supports the idea that country status is often associated with healthcare infrastructure, education, and income plays a critical role in determining life expectancy.

```{r}
#Top 10 countries by average life expectancy.
data %>%
  group_by(country) %>%
  summarise(mean_life = mean(life_expectancy)) %>%
  arrange(desc(mean_life)) %>%
  slice_head(n = 10) 
```

The countries with the highest average life expectancy are predominantly developed nations, each exceeding an average life expectancy of over 81 years.

### [MODELING]{.underline}

```{r}
#Linear Model
model<- lm(life_expectancy ~ alcohol,data)
summary(model)
```

**life_expectancy = β0 + β1 Alcohol_consumption + ε**

-   H0: **β1=0**\
    *(*There is no linear relationship between alcohol consumption and life expectancy)

-   **H1: β1\<0**\
    *(*There is a negative linear relationship between alcohol consumption and life expectancy)

**life_expectancy = 65.31 + 0.88​⋅alcohol + ε**

```{r}
# Visual checks
plot(model, which = 1)  # Linearity & homoscedasticity
plot(model, which = 2)  # Normality of residuals

# Statistical tests
shapiro.test(residuals(model))  # Normality
library(lmtest)
dwtest(model)                   # Independence of residuals

```

-   All assumptions are **not** met.

-   The residuals are not evenly spread around zero suggests **non-linearity.**

-   Residuals deviate from the normal line, especially at the tails residuals are **not normally distributed**. Reject normality.

```{r}
#confidence interval 
confint(model, level = 0.95)
```

We are 95% confident that for each 1 litre increase in alcohol consumption, life expectancy increases between 0.78 and 0.98 years, on average.

**Based on ICOR**

```{r}
#Plot highlight low icor (<0.5)
data |>
  ggplot(aes(
    x = alcohol, 
    y = life_expectancy, 
    color = income_comp))+
  geom_point( alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Life Expectancy, Alcohol, and ICOR", 
       x = "Alcohol Consumption (in litres of pure alcohol)",
       y = "Life expectancy(in age)")+
  theme_minimal()+scale_color_viridis_b()+
  gghighlight(income_comp< 0.550)

```

Among countries with low income composition (ICOR \< 0.55), higher alcohol consumption is associated with significantly lower life expectancy.

```{r}
#Plot of Alcohol, life expectancy based on ICOR
data%>%
  ggplot(aes(
    x = alcohol, 
    y = life_expectancy, 
    color = income_comp))+
  geom_point( alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Life Expectancy, Alcohol, and ICOR", 
       x = "Alcohol Consumption (in litres of pure alcohol)",
       y = "Life expectancy(in age)")+
  theme_minimal() +scale_color_viridis_b()
```

```{r}
#summary statistic of alcohol consumption and life expectancy based on ICOR
data %>%
  select(alcohol, life_expectancy,income_comp ) %>% 
  filter (income_comp>= 0.5)%>%
  summarise (mean_alcohol = mean(alcohol),
             sd_Alcohol = sd(alcohol),
             mean_life.expectancy = mean(life_expectancy),
             sd_life.expectancy = sd(life_expectancy))

data %>%
  select(alcohol, life_expectancy, income_comp) %>% 
  filter (income_comp < 0.5)%>%
  summarise (mean_alcohol = mean(alcohol),
             sd_Alcohol = sd(alcohol),
             mean_life.expectancy = mean(life_expectancy),
             sd_life.expectancy = sd(life_expectancy))

```

Alcohol consumption and life expectancy is higher in high ICOR countries.

```{r}
# Create a new variable for income group
data$income_group <- ifelse(data$income_comp < 0.55, "< 0.55", "≥ 0.55")

# Fit interaction model
model2 <- lm(life_expectancy ~ alcohol * income_comp, data = data)
summary(model2)
```

life_expectancy=54.08−2.57⋅acohol+24.45⋅income_comp+3.49⋅(alcohol×income_comp)

```{r}
# Predict values
data$predicted <- predict(model2)

# Plot
ggplot(data, aes(x = alcohol, y = life_expectancy, color = income_group)) +
  geom_point(alpha = 0.3)

```

**COMPARE MODELS**

```{r}
anova(model, model2)
```

The interaction between alcohol and income composition is highly statistically significant, suggesting that the effect of alcohol on life expectancy depends on income level.
